Disentangling the spinal mechanisms of illusory heat and burning sensations in the Thermal Grill Illusion

### 2024-01-18 Manuscript.Rmd
final ms changes

While most previous studies have focused on the TGI as a single experience, explicitly understood as an experience of or like pain, our approach differed.
We instructed participants to report the sensation from just one of the thermal components, corresponding to the cold thermode location (Exp. 1) or the warm thermode location (Exp. 2).
Importantly, the aim of this method was not to isolate any possible independent effect of warm and cold stimulation during TGI, but rather to ensure consistency in participants' ratings across trials and different stimulus conditions.To investigate both the thermosensory and burning components in each experiment, we asked participants to provide a subjective evaluation of multiple sensory qualities - cold, warmth and burning (pain).
Stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold (Exp. 1) or a warm (Exp. 2) stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli).
All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments (Figure 1). This arrangement of cold and warm probes on the skin was specifically chosen to be able to explore the spinal mechanisms of the TGI and similar set-ups, using two probes to induce TGI, have been used previously in healthy human participants [@fardo_organization_2018; @defrin_spatial_2008].

@@

We calibrated TGI stimuli by identifying a cold-warm temperature pair based on specific criteria: (1) consistently eliciting a burning sensation of at least 15 on a scale ranging from 0 to 100, (2) consistently avoiding a burning sensation (less than 15) when the cold-neutral (Exp. 1) or warm-neutral (Exp. 2) stimuli were presented, (3) both cold and warm temperatures falling within the innocuous range based on individual cold and heat pain thresholds. TGI temperature pairs, starting at 25ºC and 35ºC were presented on the participant’s forearm for 10 seconds. After 10 seconds, the participant had to rate the perceived burning (in a VAS scale from 0 - 100) coming from either the cold (experiment 1) or warm (experiment 2) probe. If the participant did not rate their perceived burning as above 15 on the scale, the experimenter increased the temperature on the warm probe, and decreased the temperature on the cold probe systematically. They then placed the probes at a different location on the forearm and repeated the trial. If the participant rated above 15ºC, the temperature combination was repeated another six times. To determine the TGI temperatures, the participant needed to rate the probe as above 15 on the burning VAS four times out of six. A list of temperatures used to calibrate the TGI stimuli can be found in the Supplementary Materials. In situations where the participant did not consistently report burning for temperatures that were below their pain thresholds, the maximum and minimum possible temperatures were used (2ºC below their thresholds).
For pain threshold measurements and TGI calibration, we positioned the probes within a single dermatome.

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 3: Thermosensory and burning components of TGI are spinally mediated: The placement of warm and cold stimuli in different dermatomes reduces the experience of the TGI.  The difference between within and across dermatome conditions are displayed for each type of stimulation (Non-TGI and TGI) for each VAS rating quality (cold, warm and burning). Positive values represent higher ratings within dermatomes, negative values represent higher ratings across dermatomes. In experiment one (A), where participants judged sensations at the location of the cold thermode, cold ratings were significantly reduced during TGI when thermodes were placed across dermatomes. In experiment two (B), where participants judged sensations at the location of the warm thermode, warm and burning ratings were significantly reduced during TGI. Small dots are individual subject means, large dots are population means for each condition and error bars are 95% confidence intervals. Please note that this figure does not include trial data where VAS = 0, this is to reflect the outcome of our statistical model."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 4: Proximodistal bias in cold perception: The spatial location of the warm and cold thermodes within dermatomes affects the perception of cold. The difference between the proximal and distal location of the reference thermode in the within dermatome are displayed by condition and by stimulation type (Non-TGI, TGI) for all VAS rating types (cold, warm, burn). Higher values show higher ratings when the cold probe is more proximal than the warm probe. In experiment one (A), where participants judged sensations at the location of the cold thermode, cold and burning ratings were higher when the cold probe was more distal. In experiment two (B), where participants judged sensations at the location of the warm thermode, cold ratings were higher when the cold probe was more distal. Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals. Please note that this figure does not include trial data where VAS =0, this is to reflect the outcome of our statistical model."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 5: Directional effects of inter-segmental sensory integration: Spatial location of warm and cold thermodes across dermatomes affects the perception of TGI and non-TGI stimuli. The difference between the caudal and rostral dermatomes are displayed by conditions and stimulation type (Non-TGI, TGI) for all VAS rating types (cold, warm, burn). Higher values represent higher ratings when the cold probe is located in dermatomes that are related to more caudal spinal segments (T1) than the warm probe (C6). In experiment one (A), where participants judged sensations at the location of the cold thermode, perception of warm and cold, cold ratings were reduced and burning ratings increased when the cold probe was located in more caudal dermatomes. In experiment two (B), where participants judged sensations at the location of the warm thermode, the changes in warm and cold ratings were similar to experiment one but specific to TGI stimuli. Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals. Please note that this figure does not include trial data where VAS =0, this is to reflect the outcome of our statistical model."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 6: Schematic of the spinal neuroanatomy directly associated with warm and cold afferents (A) within the same spinal segment and (B) across two separate spinal segments. The Lissaur’s tract is depicted as the peripheral warm and cold afferents that span to neighbouring spinal segments. Propriospinal projections from the stimulated segments possibly inhibit responses in neighbouring segments, which means integration of warm and cold sensory signals is more likely within (A) than across (B) dermatomes. The black arrow indicates direction of supraspinal structures such as the thalamus."}
ggdraw() +
    draw_image(magick::image_read(file.path("figures","figure6.png")), scale = 1)
```

@@

We thank Małgorzata Basińska, Patrik Molnár, Sára A. Szabó for their help with participant recruitment and data collection.
We also thank Michael Clements for the collection of pilot data that was instrumental in refining the experimental design.
This study was supported by a European Research Council Starting Grant (ERC-2020-StG-948788).

@@

# Supplementary methods
For the TGI calibration procedure, the experimenter followed a set list of cold-warm temperature pairs, at which to test TGI:

1. 27-35ºC
2. 25-37ºC
3. 22-38ºC
4. 20-40ºC
5. 19-41ºC
6. 18-42ºC
7. 17-43ºC
8. 16-44ºC

The experimenter initially tested the TGI using the first temperature combination. If this temperature combination did not produce a burning sensation (above 15/100 on the burning VAS) then the experimenter moved to the second temperature combination. The experimenter proceeded down the above list like so until one of two things occurred. (1) The participant responded that the temperature combination produced burning (15 on the burning VAS scale) in at least 4/6 trials or (2) temperatures were within 1-2ºC of the participant’s pain thresholds. If (1), the experimenter proceeded to the main experiment. If (2), the experimenter deviated from the above list to identify a temperature combination that produced TGI whilst remaining within the innocuous range for that participant. If the end temperature combination was reached (16-44) and the participant still did not feel TGI, then the temperature of the cold probe was lowered gradually, until either the participant experienced the TGI or the probe was within 1-2ºC of the noxious threshold.

# Supplementary results
Tables of main effects and interactions for each sensory quality (cold, warm, burning) for experiment 1 and experiment 2.

>>>>

While most previous studies have focused on the TGI as a single experience, explicitly understood as an experience of or like pain, our approach differed.
We instructed participants to report the sensation from just one of the thermal components, corresponding to the cold thermode location (Exp. 1) or the warm thermode location (Exp. 2).
Importantly, the aim of this method was not to isolate any possible independent effect of warm and cold stimulation during TGI, but rather to ensure consistency in participants' ratings across trials and different stimulus conditions.To investigate both the thermosensory and burning components in each experiment, we asked participants to provide a subjective evaluation of multiple sensory qualities - cold, warmth and burning (pain).
Stimuli consisted of either cold-warm pairs (TGI stimuli), which potentially evoked an illusion of heat and pain, or control stimuli that involved pairing a cold (Exp. 1) or a warm (Exp. 2) stimulus of identical temperature as in the TGI condition with a baseline temperature of 30°C (non-TGI stimuli).
All stimulation pairs were presented at a fixed distance on the skin, either within the same dermatome or across dermatomes that mapped onto non-adjacent spinal segments (Fig. 1). This arrangement of cold and warm probes on the skin was specifically chosen to be able to explore the spinal mechanisms of the TGI and similar set-ups, using two probes to induce TGI, have been used previously in healthy human participants [@fardo_organization_2018; @defrin_spatial_2008].

@@

We calibrated TGI stimuli by identifying a cold-warm temperature pair based on specific criteria: (1) consistently eliciting a burning sensation of at least 15 on a scale ranging from 0 to 100, (2) consistently avoiding a burning sensation (less than 15) when the cold-neutral (Exp. 1) or warm-neutral (Exp. 2) stimuli were presented, (3) both cold and warm temperatures falling within the innocuous range based on individual cold and heat pain thresholds. TGI temperature pairs, starting at 25ºC and 35ºC were presented on the participant’s forearm for 10 seconds. After 10 seconds, the participant had to rate the perceived burning (in a VAS scale from 0 - 100) coming from either the cold (Exp. 1) or warm Exp. 2) probe. If the participant did not rate their perceived burning as above 15 on the scale, the experimenter increased the temperature on the warm probe, and decreased the temperature on the cold probe systematically. They then placed the probes at a different location on the forearm and repeated the trial. If the participant rated above 15 on the VAS, the temperature combination was repeated another six times. To determine the TGI temperatures, the participant needed to rate the probe as above 15 on the burning VAS four times out of six. In situations where the participant did not consistently report burning for temperatures that were below their pain thresholds, the maximum and minimum possible temperatures were used (2ºC below their thresholds).
For pain threshold measurements and TGI calibration, we positioned the probes within a single dermatome.

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 3: Thermosensory and burning components of TGI are spinally mediated: The placement of warm and cold stimuli in different dermatomes reduces the experience of the TGI.  The difference between within and across dermatome conditions are displayed for each type of stimulation (Non-TGI and TGI) for each VAS rating quality (cold, warm and burning). Positive values represent higher ratings within dermatomes, negative values represent higher ratings across dermatomes. In experiment one (A), where participants judged sensations at the location of the cold thermode, cold ratings were significantly reduced during TGI when thermodes were placed across dermatomes. In experiment two (B), where participants judged sensations at the location of the warm thermode, warm and burning ratings were significantly reduced during TGI. Small dots are individual subject means, large dots are population means for each condition and error bars are 95% confidence intervals. Please note that to best reflect the outcome of our zero-inflated statistical model this figure only includes trial data where VAS ratings were greater than 0."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 4: Proximodistal bias in cold perception: The spatial location of the warm and cold thermodes within dermatomes affects the perception of cold. The difference between the proximal and distal location of the reference thermode in the within dermatome are displayed by condition and by stimulation type (Non-TGI, TGI) for all VAS rating types (cold, warm, burn). Higher values show higher ratings when the cold probe is more proximal than the warm probe. In experiment one (A), where participants judged sensations at the location of the cold thermode, cold and burning ratings were higher when the cold probe was more distal. In experiment two (B), where participants judged sensations at the location of the warm thermode, cold ratings were higher when the cold probe was more distal. Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals. Please note that to best reflect the outcome of our zero-inflated statistical model this figure only includes trial data where VAS ratings were greater than 0."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 5: Directional effects of inter-segmental sensory integration: Spatial location of warm and cold thermodes across dermatomes affects the perception of TGI and non-TGI stimuli. The difference between the caudal and rostral dermatomes are displayed by conditions and stimulation type (Non-TGI, TGI) for all VAS rating types (cold, warm, burn). Higher values represent higher ratings when the cold probe is located in dermatomes that are related to more caudal spinal segments (T1) than the warm probe (C6). In experiment one (A), where participants judged sensations at the location of the cold thermode, perception of warm and cold, cold ratings were reduced and burning ratings increased when the cold probe was located in more caudal dermatomes. In experiment two (B), where participants judged sensations at the location of the warm thermode, the changes in warm and cold ratings were similar to experiment one but specific to TGI stimuli. Small points show data from each participant, large dots are means across trials for each condition, and error bars show 95% confidence intervals. Please note that to best reflect the outcome of our zero-inflated statistical model this figure only includes trial data where VAS ratings were greater than 0."}

@@

```{r, results="hide", fig.width = 7.2, fig.height = 5, warning = F, message = F, echo = F, fig.cap = "Figure 6: Schematic of the spinal neuroanatomy associated with warm and cold afferents (A) within the same spinal segment and (B) across two separate non-adjacent spinal segments. After entering the spinal cord, peripheral warm and cold afferent fibres form collateral branches and synapses above and below the level of entry in neighbouring spinal segments. Propriospinal neurons possibly inhibit responses in neighbouring segments (i.e., inter-segmental inhibition), which means integration of warm and cold sensory signals is more likely within (A) than across (B) dermatomes. Projection neurons decussate and ascend contralaterally forming the spinothalamic pathway. The black arrow indicates direction of supraspinal structures such as the thalamus."}
ggdraw() +
    draw_image(magick::image_read(file.path("figures","figure6.png")), scale = 1)
```

@@

We thank Małgorzata Basińska, Patrik Molnár, Sára A. Szabó for their help with participant recruitment and data collection.
We also thank Michael Clements for the collection of pilot data that was instrumental in refining the experimental design.
This study was supported by a European Research Council Starting Grant (ERC-2020-StG-948838).

@@

# Supplementary materials
Tables of main effects and interactions for each sensory quality (cold, warm, burning) for experiment 1 and experiment 2.
